---
title: "Untitled"
author: "Brian Boughton"
date: "2023-04-25"
output: html_document
---

# Milestone 5 -Naive Bayes

## Goal

Implement NB model with 80-20 and 60-40 split evaluation. If you implement 4-fold cross-validation analysis, you will get 10-point bonus.


* another classification model with training-test split of 80-20 and 60- 40 and results of precision, recall, and f-measure. (Note, if you work as a group, you need to pick another algorithm different from your Milestone 2 with another feature set after the feature selection.)

* A model on 4-fold cross-validation with the results of precision, recall, and f-measure. (Bonus)

* a report to summarize your finding. It may include but not limit to the questions, such as which model performances better in this dataset and which split performs better? Why?

```{r}
dt<-read.csv("hour.csv")
dt<-subset(dt, select = -c(instant, dteday,casual,registered,yr))

dt$temp<-ifelse(dt$temp >= mean(dt$temp), 1, 0)
dt$atemp<-ifelse(dt$atemp >= mean(dt$atemp), 1, 0)
dt$hum<-ifelse(dt$hum >= mean(dt$hum), 1, 0)
dt$windspeed<-ifelse(dt$windspeed >= mean(dt$windspeed), 1, 0)
dt$weathersit<-ifelse(dt$weathersit == 4,3,dt$weathersit)
dt$cnt<-ifelse(dt$cnt >= mean(dt$cnt), 1, 0)
dt[,1]<-as.factor(dt[,1])
dt[,2]<-as.factor(dt[,2])
dt[,3]<-as.factor(dt[,3])
dt[,4]<-as.factor(dt[,4])
dt[,5]<-as.factor(dt[,5])
dt[,6]<-as.factor(dt[,6])
dt[,7]<-as.factor(dt[,7])
dt[,8]<-as.factor(dt[,8])
dt[,9]<-as.factor(dt[,9])
dt[,10]<-as.factor(dt[,10])
dt[,11]<-as.factor(dt[,11])
dt[,12]<-as.factor(dt[,12])

```

```{r}
train.index <- sample(c(1:dim(dt)[1]), dim(dt)[1]*0.6)  
train.df <- dt[train.index, ]
valid.df <- dt[-train.index, ]
```

```{r}
(nb<-naiveBayes(cnt~., data=train.df))
```

```{r}
pred.prob <- predict(nb, newdata = valid.df, type = "raw")
pred.class <- ifelse(pred.prob[,1]>0.5, 1, 0)
```

```{r}
library(caret)

# training
pred.training.prob <- predict(nb, newdata = train.df, type="raw")
pred.training.class <- ifelse(pred.training.prob[,1]>0.5, 0, 1)
confusionMatrix(as.factor(pred.training.class),as.factor(train.df$cnt))
```

```{r}
# validation
pred.prob <- predict(nb, newdata = valid.df, type="raw")
pred.class <- ifelse(pred.prob[,1]>0.5, 0, 1)
confusionMatrix(as.factor(pred.class), as.factor(valid.df$cnt))
```



